### 2020-02-10 doc/icfp20/counterpoint.tex
add figures & start second species section

@@

Counterpoint is a technique for combining multiple lines of melodies.
Composing counterpoint is like arranging a song for choir:
we start with a \emph{cantus firmus}, which serves as the base melody,
and compose counterpoint lines above or below the cantus firmus.
When doing this, we must make sure that the counterpoint lines sound
harmonically pleasing when played with the cantus firmus,
and that the individual melodic lines are distinguishable to the listener.

@@

First species counterpoint is the simplest variant of counterpoint.
In first species, we set one note against each note in the cantus firmus,
which starts with a tonic and consists only of whole notes.
Figure~\ref{fig:frog} is an example of first species counterpoint.
The lower line is the cantus firmus, which we take from a famous
German song called ``Froschgesang'' (Frog's song).
The upper line is the counterpoint composed by the second author.

@@

In our implementation, we represent each bar as a pitch-interval
pair \texttt{(p ,  i)}, where \texttt{p} is a note in the cantus firmus,
and \texttt{i} is the distance between \texttt{p} and the corresponding
counterpoint note.
We then represent the sequence of bars as a list of pitch-interval pairs,
but with one proviso: we separate the first and last bars from the
middle bars.
Therefore, the Frog's song in Figure~\ref{fig:frog} is represented as a
compound of the following three elements:

@@

The reason why we separate different parts of music is that
they are subject to different rules.
Below, we elaborate these rules one by one.

@@

The beginning of the music should express perfection.
Therefore, the first interval must be a unison, a 5th, or an 8th.
In our formalization, we implement this rule as a function
\texttt{checkBeginning}, which reports an error \texttt{not158 i}
when the first interval \texttt{i} is an invalid one.

@@

@@

>>>>

\newcommand{\FS}{
  \begin{figure}
    \includegraphics[width=10cm]{fs.png}
    \caption{First Species Counterpoint}
    \label{fig:fs}
  \end{figure}
}

\newcommand{\SS}{
  \begin{figure}
    \includegraphics[width=10cm]{ss.png}
    \caption{Second Species Counterpoint}
    \label{fig:ss}
  \end{figure}
}

@@

Counterpoint is a technique for combining multiple lines of melodies.
Composing counterpoint is like arranging a song for choir:
we start with a \emph{cantus firmus}, which serves as the base melody,
and compose counterpoint lines above or below the cantus firmus.
When doing this, we must make sure that the whole music sounds
harmonically pleasing, and that the individual melodic lines are
distinguishable to the listener.

@@

\FS

First species counterpoint is the simplest variant of counterpoint.
In first species, we set one note against each note in the cantus firmus,
which starts with a tonic and consists only of whole notes.
Figure~\ref{fig:fs} is an example of first species counterpoint.
The lower line is the cantus firmus, which we take from a famous
German song called ``Froschgesang'' (Frog's song).
The upper line is the counterpoint composed by the second author.

@@

In our implementation, we represent each bar as a pitch-interval
pair \texttt{(p ,  i)}, where \texttt{p} is a note in the cantus firmus,
and \texttt{i} is the distance between \texttt{p} and the corresponding
counterpoint note.
We then represent a sequence of bars as a list of pitch-interval pairs,
but with one proviso: we separate the first and last bars from the
middle bars.
Therefore, the Frog's song in Figure~\ref{fig:fs} is represented as a
compound of the following three elements:

@@

The reason behind our three-part representation of music is that
different parts are subject to different rules.
Let us go through the

@@

The beginning of the music should express perfection.
Therefore, the first interval must be a unison, a 5th, or an 8th.
In our formalization, we implement this rule as a function
\texttt{checkBeginning}, which reports an error \texttt{not158 i}
when the first interval \texttt{i} is an invalid one.
Since Agda does not have exceptions, we wrap this error around
the \texttt{just} constructor for option values.

@@

\noindent Recall that the functions representing these rules return
either an option value or a list of found errors.
Therefore, the correctness proofs either take the form
\texttt{func args $\equiv$ nothing}, or look like \texttt{func args $\equiv$ []}.

@@

\subsection{Second Species Counterpoint}
\label{sec:cp:ss}

\SS

We next discuss a more complex variant of counterpoint, called the
second species.
In second species, we set \emph{two} half notes against every cantus
firmus note.
Figure~\ref{fig:ss} shows an example of such two-against-one
counterpoint, again composed for the Frog's song.
To represent second species counterpoint, we define a new type
\texttt{PitchInterval2} as a triple of a pitch and two intervals:

\begin{alltt}
PitchInterval2 : Set
PitchInterval2 = Pitch \(\times\) Interval \(\times\) Interval
\end{alltt}

The rhythmic structure of secound species counterpoint gives rise to
the distinction between \emph{strong beats} (the first interval in a bar)
and \emph{weak beats} (the second interval).
Roughly speaking, strong beats in middle bars are constrained by the
same rules as in first species.
That is, they must all be consonant, non-unison intervals, and in the 
case of perfect intervals, the motion from the preceding interval (on
the weak beat) must be either contrary or oblique.
Weak beats, on the other hand, ...
